---
title: "Mixed Manova on Collapased Sample-Time Data"
author: "Euphy Wu"
date: "2024-10-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this illustration is to demonstrate the use of current
packages for implementing MANOVA with random effects or multivariate
generalized mixed effects models. The response variables are transformed
cell type proportions for each sample at a given timepoint, while the
fixed variable is `Time` only, with random effects accounting for
repeated measures within each sample.

The following two packages are being considered:

-   `"MANOVA.RM`

-   `MCMCglmm`

`MCMCglmm` fits a multivariate linear mixed effects model using MCMC,
while `MANOVA.RM` fits a MANOVA model with repeated measures.

Currently, the `MMeM` package only supports multivariate mixed effects
models with one fixed effect, one random effect, and two response
variables. Since we typically have more than two response variables
(proportions), `MMeM` is not suitable for this illustration.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide', include = FALSE}
library(Matrix)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(cowplot)
require(pals)
library(MASS)
library(tibble)
library(MANOVA.RM)
library(MCMCglmm)
setwd("/proj/milovelab/wu/scLDAseq")
r.file <- paste0("R/",list.files("R/"))
sapply(r.file, source)
dir <- "/work/users/e/u/euphyw/scLDAseq/data/simulation/MultiSample_LogisticNormal/NormalGamma_SingleResponse/nSample20_nCellType5_noBatch_CancerCell/"
file_name <- "scSTM_1728956995_HighVar.rds"
set_level <- sub("^scSTM_(.*)\\.rds$", "\\1", file_name)
scSTM_name <- paste0(dir, "scSTM_LinearMixed_noContent_Prevalence_Time/scSTM_",
                     set_level,".rds")
```

First, we read in an example `scSTMseq` object under the alternative
hypothesis, which assumes that there are proportion changes between
timepoints.

```{r}
if(file.exists(scSTM_name)){
  scSTMobj <- readRDS(scSTM_name)
  scSTMobj <- select_top_scSTM(scSTMobj)
}else{
  next
}
scSTMobj
```

In the following implementation, we ignore the posterior replicates and
focus on a single replicate. Here are the steps for transforming the
model output into response variables used in fitting MANOVA or GLMM
models:

1.  extract $\theta$ from the model which is topic proportion for each
    cell
2.  collapse the proportion by sample and time
3.  re-normalize the data so that the cell type proportion sums to 1 for
    each sample/time combination $\theta^*$

```{r}
# step 1: extract theta
theta.old <- scSTMobj$theta

# step 2: collapse by time and sample
metadata <- colData(scSTMobj$settings$sce) %>% as.data.frame()
theta.old <- theta.old %>%
      as.data.frame() %>%
      rownames_to_column("Cell") %>%
      left_join(metadata, by = "Cell")

theta.collapsed <- theta.old %>%
  group_by(Sample, Time) %>%
  summarise(across(starts_with("V"), sum), .groups = 'drop') %>%
  ungroup()

# step 3: renormalization
theta.new <- theta.collapsed %>%
  rowwise() %>%
  mutate(across(starts_with("V"), ~ . / sum(c_across(starts_with("V"))))) %>%
  ungroup()

head(theta.new)
```

Before using `theta.new` as the response variable, we apply the
isometric log-ratio transformation.

```{r}
theta.irl <- theta.new %>% 
  dplyr::select(starts_with("V"))
theta.irl <- compositions::ilr(theta.irl)
res <- theta.new %>% 
  dplyr::select(-starts_with("V"))
theta.irl <- cbind(res, theta.irl) %>%
  as.data.frame()%>%
  dplyr::mutate(Time = as.factor(Time),
         Sample = as.factor(Sample))
head(theta.irl)
```

## `MANOVA.RM`

-   CRAN: <https://cran.r-project.org/web/packages/MANOVA.RM/index.html>

-   Vignettes:
    <https://cran.r-project.org/web/packages/MANOVA.RM/vignettes/Introduction_to_MANOVA.RM.html>

### Toy Data

Let's start with a simple simulation to demonstrate repeated measures
analysis using `MANOVA.RM`. Assume we have 6 patients, each with
`Weights` measured at two timepoints. We also create an unrelated
covariate `Grou`p.

```{r}
n_patients <- 6
patients <- factor(rep(1:n_patients, each = 2))
time <- rep(c(0, 1), n_patients)
# Random intercept for each patient
random_intercept <- rnorm(n_patients, mean = 70, sd = 10) 
# assume weight increases by 2 unit, at time = 2
weight <- random_intercept[as.numeric(patients)] + 2 * time + rnorm(n_patients * 2, sd = 1)
Group <- rep(1:2, each = n_patients)
data <- data.frame(Patient = patients, Time = time, Weight = weight, Group = Group)
```

We fit `RM` to assess the change in weight, considering repeated
measures within patients.

```{r}
fit.rm <- RM(Weight ~ Group + Time, data = data, subject = "Patient", no.subf = 1, iter = 1000)
summary(fit.rm)
```

For comparison, we fit a simple linear mixed effects model using the
same data.

```{r}
lm.mixed <- lmer(Weight ~ (1|Patient) + Time + Group, data = data)
summary(lm.mixed)
```

The results for $H_0$: Time Effect = 0 are consistent between the two
models, while the unrelated covariate `Group` is rejected.

### Cell Data

Next, we apply MANOVA with repeated measures to our single-cell data,
testing if the time effect equals 0.

```{r}
response_vars <- grep("^V", names(theta.irl), value = TRUE)
fixed_formula <- as.formula(paste0("cbind(", paste(response_vars, collapse = ", "), ") ~ Time"))
start_time <- proc.time()
fit.temp <- multRM(fixed_formula, data = theta.irl, subject = "Sample", within = "Time", iter = 1000)
end_time <- proc.time()
fit.temp
```

We can also check the computation time for `multRM.`

```{r}
elapsed_time <- end_time - start_time
print(elapsed_time)
```

We evaluate the assumption of homogeneity of variance. Since the
covariance matrix is non-singular, both WTS and MATS tests are valid.

```{r}
 # check for singular covariance matrix
  test <- try(solve(fit.temp$Covariance), silent = TRUE)
  if(!is.matrix(test)){
    warning("The covariance matrix is singular. The WTS provides no valid test statistic!")
  }
```

## MCMCglmm

-   Tutorial: <https://github.com/jarrodhadfield/MCMCglmm>

-   CRAN:<https://cran.r-project.org/web/packages/MCMCglmm/index.html>

The code follows Chapter 5 of course note:
<https://github.com/jarrodhadfield/MCMCglmm/blob/master/vignettes/CourseNotes.pdf>

### Toy Data

We first fit the toy data created before, where `Time` and `Group` are
the fix effects, and `Patient` is the random effect. The prior is a flat
improper prior. We can see that the `Time` estimate was close to 2, with
the 95% HBD includes the true value $2$. However, the `Group` effect was
boarderline significant, while it is completely random.

```{r}
prior = list(R = list(V = 1, nu = 0.002), G = list(G1 = list(V = 1,nu = 0.002)))
mcmc.fit.toy <- MCMCglmm(Weight ~ Time + Group, random = ~Patient, data = data, 
                     verbose = FALSE, prior = prior)
summary(mcmc.fit.toy)
```

### Cell Data

We then implement `MCMCglmm` on our `scSTMseq` data, with improper
priors. In the output, `traitV1` is equivalent to `topic 1` , same for
`traitV2`-`traitV4`.

```{r }
response_vars <- grep("^V", names(theta.irl), value = TRUE)
fixed_formula <- as.formula(paste0("cbind(", paste(response_vars, collapse = ", "), ") ~ Time "))
fixed_formula
priors = list(R = list(V = diag(4), nu = 0.002),
             G = list(G1 = list(V = diag(4), nu = 2,
                                alpha.mu = rep(0, 4),
                                alpha.V = diag(1, 4, 4))))
start_time <- proc.time()
mcmc.fit <- MCMCglmm(fixed_formula, random = ~us(trait):Sample, rcov = ~us(trait):units, data = theta.irl, family = rep("gaussian", 4),
                     verbose = FALSE, prior = priors)
end_time <- proc.time()
summary(mcmc.fit)
```

We also check the run time for `MCMCglmm`

```{r}
elapsed_time <- end_time - start_time
print(elapsed_time)
```

## Conclusion

It is challenging to identify which model is better based on one
example, so the next steps are

1.  Detail the methods behind each model.

2.  Scale up the implementation to assess Type I error and power.
