---
title: "figure_on_real_data"
output: html_document
date: "2025-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we read in the data. Since we had 100 posterior replicate for each sample at a given timepoint, we can get the final cell type proportion by taking the mean of the 100 posterior replicates. Note the proportion is averaged across all samples. 
```{r, warning=FALSE, message=FALSE}
setwd("/proj/milovelab/wu/scLDAseq")
library(Matrix)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyverse)

collaposed.theta <- readRDS( "res/PD1/collaposed_theta.rds")
############### proportion change by E and NE ##############
average_by_group <- lapply(collaposed.theta, function(rep){
  avg_group <- rep %>%
    group_by(expansion, timepoint) %>%
    summarise(across(starts_with("V"), mean, .names = "mean_{.col}"), .groups = "drop")
})
average_by_group <- bind_rows(average_by_group)
average_by_group_long <- average_by_group %>%
  pivot_longer(cols = starts_with("mean"), names_to = "CellType", values_to = "Proportion")

average_by_group_long <- average_by_group_long %>%
  mutate(CellType = gsub("mean_V", "Group", CellType))
```
```{r}
head(average_by_group_long)
```

As request, the x axis represent times, and the y-axis are the proportions from each topic and timepoint
```{r}
ggplot(average_by_group_long, aes(x = timepoint, y = Proportion, color = CellType)) +
    geom_boxplot(aes(fill = CellType), alpha = 0.3, outlier.shape = NA) +
    labs(
        title = "Proportion of Groups by Timepoint",
        x = "Timepoint",
        y = "Proportion",
        color = "Group",
        fill = "Group"
    ) +
    theme_minimal(base_size = 14) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1")

```

For better visualization, I have changed the x-axis to cell groups and colored by time

```{r}
ggplot(average_by_group_long, aes(x = CellType, y = Proportion, color = timepoint)) +
    geom_boxplot(aes(fill = timepoint), alpha = 0.3, outlier.shape = NA) +
    labs(
        title = "Proportion of Groups by Timepoint",
        x = "CellType",
        y = "Proportion",
        color = "Time",
        fill = "Time"
    ) +
    theme_minimal(base_size = 14) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1")
```

Repeat the plots to have the proportion on inverse logit $exp(y)/(1+exp(y))$

```{r}
ggplot(average_by_group_long, aes(x = timepoint, y = exp(Proportion)/(1+exp(Proportion)), color = CellType)) +
    geom_boxplot(aes(fill = CellType), alpha = 0.3, outlier.shape = NA) +
    labs(
        title = "Proportion of Groups by Group",
        x = "Timepoint",
        y = "Inverse logit Proportion",
        color = "Group",
        fill = "Group"
    ) +
    theme_minimal(base_size = 14) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1")
```

```{r}
ggplot(average_by_group_long, aes(x = CellType, y = exp(Proportion)/(1+exp(Proportion)), color = timepoint)) +
    geom_boxplot(aes(fill = timepoint), alpha = 0.3, outlier.shape = NA) +
    labs(
        title = "Proportion of Groups by Timepoint",
        x = "CellType",
        y = "Inverse logit Proportion",
        color = "timepoint",
        fill = "timepoint"
    ) +
    theme_minimal(base_size = 14) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1")

```

