---
title: "propeller_benchmark"
output: html_document
date: "2024-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Apply Propeller to Seurat Output
```{r,warning=F, message=F}
library(Matrix)
library(dplyr)
library(scuttle)
library(tidyverse)
library(SingleCellExperiment)
library(MASS)
library(speckle)
library(limma)
library(CellBench)
```

Since `seurat` was run with multiple resolution, the resolution with the highest clustering accuracy was selected for downstream analysis
```{r}
setwd("/proj/milovelab/wu/scLDAseq")
seurat_name <- "/work/users/e/u/euphyw/scLDAseq/data/simulation/1MultiSample/SingleResponse/nSample20_nCellType10_noBatch_StromalCell/seurat/seurat_1730958979_HighVar0.16.rds"
sims_name <- "/work/users/e/u/euphyw/scLDAseq/data/simulation/1MultiSample/SingleResponse/nSample20_nCellType10_noBatch_StromalCell/sims/sims_1730958979_HighVar0.16.rds"
sims <- readRDS(sims_name)
seurat.sims <- readRDS(seurat_name)
smeta <- seurat.sims@meta.data %>% as.data.frame()
sub_sims <- sims[,rownames(smeta)] # filter by the rows
seurat.adj <- sapply(smeta[,4:7], function(x) {
  adjustedRandIndex(x, sub_sims$Group)
})
# select the resolution that has the highest ARI. When there are multiple, select the first one
best_res <- names(seurat.adj)[seurat.adj == max(seurat.adj)][1]
seurat_cluster <- seurat.sims@meta.data %>% as.data.frame() %>% dplyr::select(all_of(best_res))
Idents(seurat.sims) = seurat_cluster$RNA_snn_res.0.5
seurat.sims$Time <- colData(sims)$Time[match(rownames(seurat.sims@meta.data),colData(sims)$Cell)]
seurat.sims$Sample <- colData(sims)$Sample[match(rownames(seurat.sims@meta.data),colData(sims)$Cell)]
```

Collapose the clustering information to clustering proportion by cell type and sample, for each timepoint. This follows propeller code, other than in propeller, they do not account for timepoints.
```{r}
clusters <- Idents(seurat.sims)
sample <- seurat.sims$Sample
group <- seurat.sims$Time

transform="logit"
trend=FALSE
robust=TRUE
tab <- table(sample, clusters, group)
props <- apply(tab, c(1, 3), function(x) x / sum(x))
props <- aperm(props, c(2, 1, 3)) 
props
```

Applying logit transformation to the proportion data for each timepoint
```{r}
message("Performing logit transformation of proportions")
props.pseudo <- apply(tab, c(1, 3), function(x) (x + 0.5) / sum(x + 0.5))
props.pseudo <- aperm(props.pseudo, c(2, 1, 3)) 
prop.trans <- log(props.pseudo/(1-props.pseudo))
prop.trans <- rbind(prop.trans[, , 1], prop.trans[, , 2])
sample <- rownames(prop.trans)
colnames(prop.trans) <- paste0("Cluster", 1:ncol(prop.trans))
prop.trans <- t(prop.trans)
```

Apply propeller to the clustering results, with sample blocks to account for random effect in samples
```{r}
group <- rep(c(0, 1), each = length(unique(sample)))
design<-model.matrix(~0+factor(group))
colnames(design) <- c("Time1", "Time2")
corfit<-duplicateCorrelation(prop.trans,design,block=sample)
fit<-lmFit(prop.trans,design, block=sample, correlation=corfit$consensus)
contrasts <- makeContrasts(Time = Time2 - Time1, levels = design)
fit.cont <- contrasts.fit(fit, contrasts = contrasts) 
fit.cont <- eBayes(fit.cont, robust=robust, trend=trend)
out <- as.data.frame(fit.cont)
rownames(out) <- rownames(fit.cont$coefficients)
out$fdr <- p.adjust(out$p.value, method="BH")
out
```

## Apply Propeller to Seurat Output
