---
title: "Simulation_Demonstration"
author: "Euphy Wu"
date: "2024-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Recap of the scSTMseq Model

Through scSTMseq, we output $\theta$, which is a N by K, cell by topic proportion matrix, where N is the number of cells, and K is the number of topics.

$\theta \sim$ LogisticNormal$(\mu_k + \psi_k, \Sigma^{(t)})$ where $k = 1,\dots, K-1$. This is equivalent to $\eta \sim$ Normal$(\mu_k + \psi_k, \Sigma^{(t)})$. $\eta_{n,K}$ is fixed at 0 for identifiability of the model for any $n = 1, \dots, N$.

$\eta_k = \gamma_{0,k} + \gamma_{1:P,k}^T X^T + \psi_k + \epsilon$ where $\psi_k$ is viewed as a sample random effect following a multivariate normal distribution with mean $0$ and standard variance. $k = 1, \dots, K-1$ in this mixed effect model.

In this simulation, we will

1)  Setup the design matrix $X$

2)  Simulate Effect Matrix $\Gamma$

3)  Simulate Sample Random Effect $\psi$

4)  Generate $\eta$ and then map to $\theta$, which is cell proportion by patients that will be fed into revised Splatter

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(Matrix)
library(dplyr)
library(scuttle)
library(tidyverse)
library("scater")
library(SingleCellExperiment)
library(MASS)
library(VariantAnnotation)
library(checkmate)
library(MCMCpack)
```

### Generate Design Matrix

The desired covariates include Timepoint, Response, and the interaction term between Timepoint and Response

```{r}
nSample <- 6
nTimepoints <- 2

sample_ids <- rep(1:nSample, each = nTimepoints) 
Timepoint <- rep(c(0, 1), nSample)
Response <- rep(c(0, 1), each = (nSample * nTimepoints) / 2)
# interaction Term
Timepoint_Response <- Timepoint * Response

# Create design matrix for each sample/time combination
X <- data.frame(Timepoint = Timepoint,
                Response = Response,
                Timepoint_Response = Timepoint_Response)
rownames(X) <- paste(paste0("Sample_", sample_ids), ifelse(Response == 0, "nonResponse", "Response"),
                     ifelse(Timepoint == 0, "t1", "t2"), sep = "_")
print(X)
```

### Simulate Effect Size $\Gamma$

In the following simulation, Timepoint has effect size 0, indicating that in non-response group, there will not be any cell type proportion change between timepoints. Response effect size is also 0, meaning no proportion difference between response and non-response group at timepoint 1. This can be changed to simulate different simulation settings.

For response group, we set `effectSize = 0.8`. For simplicity, k1 has a positive `effectSize`, and k2 has a negative `effectSize`, meaning that cell type 1 will increase and cell type 2 will decrease at time point 2. We can also vary this `effectSize`, as well as revise between different $k$ for a more complicated simulation.

```{r}
nCellType <- 5
Simplex <- nCellType - 1 # this is K-1
Response <- 0 # Suppose Being in Response Group has no effect on patient's baseline proportion
effectSize <- 0.8

gamma <- data.frame(K1 = c(0, Response, effectSize),
                   K2 = c(0, Response, -effectSize))
gamma <- cbind(gamma, matrix(0, nrow = 3, ncol = (Simplex - 2)))
# gamma <- rbind(1, gamma)
colnames(gamma) <- c(paste0("K", 1:ncol(gamma)))
rownames(gamma) <- c("Timepoint", "Response", "Timepoint_Response")

print(gamma)

```

We simulate $\mu$ by multiplying $\gamma$ and the design matrix $X$.

```{r}
mu <- t(t(as.matrix(gamma))  %*% t(as.matrix(X)))
print(mu)
```

### Generate Sample Random Effect

Sample random effect $\psi$ is generated from a multivariate normal distribution with 0 mean, and a diagnol 1' matrix.

```{r}
psi <- mvrnorm(n = nSample, mu = rep(0, Simplex),  Sigma= diag(rep(1, Simplex)))

# adding random effect eta = mu + psi
for (i in 1:nSample) {
  idx_t1 <- 2 * (i - 1) + 1  
  idx_t2 <- 2 * i  
  
  mu[idx_t1, ] <- mu[idx_t1, ] + psi[i, ]
  mu[idx_t2, ] <- mu[idx_t2, ] + psi[i, ]
}

print(mu)
```

### Generate $\eta$ and final proportion $\theta$

We first add $\eta_K = 0$ to every sample that was previous simulated to achieve a complete $\eta$.

```{r}
eta <- cbind(mu, 0)
colnames(eta) <- paste0("K", 1:ncol(eta))
print(eta)
```

Then we map $\eta$ back to the simplex $\theta$ using LogSumExp stabilization.

```{r}
map_to_simplx <- function(x) {
  exp(x - log(sum(exp(x))))
}

theta <- t(apply(eta, 1, map_to_simplx))
print(theta)
```

Now we are ready to feed this $\theta$ matrix into the revised Splatter algorithm. In Splatter, we use `sample` function to generate cell with cell type identity from $\theta$.

## Do we need the Dirichlet Regression?

The above simulation follows closely to our generative process in scSTMseq. I am leaning towards that we might not need Dirichlet regression anymore in the simulation. However, following what we have discussed during the meeting, $softmax(\theta) = X\gamma + \psi$ , I set up the following code.

First of all, since Dirichlet regression does not accept parameter with 0, we need to re-simulate the $\eta$. Instead of $K-1$, now $\Gamma$ needs $K$ dimensions. Same applies to $\psi$.

```{r}
nCellType <- 5
Response <- 0 # Suppose Being in Response Group has no effect on patient's baseline proportion
effectSize <- 0.8

gamma <- data.frame(K1 = c(0, Response, effectSize),
                   K2 = c(0, Response, -effectSize))
gamma <- cbind(gamma, matrix(0, nrow = 3, ncol = (nCellType - 2)))
# gamma <- rbind(1, gamma)
colnames(gamma) <- c(paste0("K", 1:ncol(gamma)))
rownames(gamma) <- c("Timepoint", "Response", "Timepoint_Response")

mu <- t(t(as.matrix(gamma))  %*% t(as.matrix(X)))

# simulate patient random effect
psi <- mvrnorm(n = nSample, mu = rep(0, nCellType),  Sigma= diag(rep(1, nCellType)))

# adding random effect eta = mu + psi
for (i in 1:nSample) {
  idx_t1 <- 2 * (i - 1) + 1  
  idx_t2 <- 2 * i  
  
  mu[idx_t1, ] <- mu[idx_t1, ] + psi[i, ]
  mu[idx_t2, ] <- mu[idx_t2, ] + psi[i, ]
}

eta <- mu
print(eta)
```

Shift $\eta$ to all positives, and we draw from a Dirichlet regression with shifted $\eta$ as parameters

```{r}
eta <- eta + abs(min(eta)) +1
theta <- t(apply(eta, 1, function(row) rdirichlet(1, row)))
print(theta)
```
